Copulas and their potential for ecology

### 2019-10-11 Paper.Rmd
added something on tropical

<!--***DAN: remianing to do:
DONE 1) Edit abstract
2) Go through the whole thing one more time
3) send it out to coauthors (done) and maybe get feedback, ESP FROM lAWRENCE
4)make sure to say something about tropical systems
DONE 5) maybe write to the editors saying this is coming soon?
DONE 6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.
DONE 7) change the box around the Cedar Creek data in the main fig
DONE 8) say in Discussion that you could have used non-ranks for the nonparam stats - this
is the stuff the Ecol Mongr referee talked about. Maybe this should go in Background instead?
9) maybe go back and read the Ecol Monogr reviews again
-->

@@

<!--bivariate summary-->
Table \ref{tab_bivar_summary} summarizes Q1 results 
for the bivariate datasets, with some details in the Appendices. The two 
variables were significantly related for all datasets 
(Table \ref{tab_bivar_summary}, row 1). A non-normal copula 
always emerged as the lowest-AIC copula 
(Table \ref{tab_bivar_summary}, row 2). 
The normal copula was a poor fit compared to the lowest-AIC copula 
(Table \ref{tab_bivar_summary}, rows 3-4), except for Cedar Creek, for which the AIC
difference between these fits was not meaningfully large.
Thus we answer the first part of Q1 in the affirmative for 3 of 4 of our bivariate datasets. 
Often, either lower- or upper-tail dependence statistics differed substantially from 0 
(Table \ref{tab_bivar_summary}, rows 7-8), and/or these statistics differed from each other (Table \ref{tab_bivar_summary}, row 9), 
helping to answer the second and third parts of Q1. Though again the model
selection results do not provide 
information on statistical significance of these differences, and are subject to the
caveat that, for some datasets, even the lowest-AIC copula was not an objectively good fit
(Table \ref{tab_bivar_summary}, rows 5 and 6), nonparametric results 
(Table \ref{tab_bivar_summary}, row 10) showed that tail associations
deviated significantly from what would be expected from a normal-copula
null model, and were also significantly asymmetric, except for the Cedar Creek data.
See Figs \ref{SM-fig_bmr_birds_nonparam}-\ref{SM-fig_cc2000_nonparam} for analogues to 
Fig. \ref{fig_soilCN_nonparam} for the bivariate datasets other than the soil C and N
data, see Tables \ref{SM-tab_bmr_birds_fit}-\ref{SM-tab_cc2000fit} for analogues to Table \ref{tab_soilCNfit},
and see Table \ref{SM-tab_3biv_npa_summary} for analogues to Table \ref{tab_soilCN_asym}.
The first three datasets had stronger upper- 
than lower-tail dependence, and the last dataset had the reverse, though non-significantly. Our 
generally affirmative (except for Cedar Creek), 
empirically based answer to Q1 for the bivariate datasets is 
represented in Fig. \ref{MasterFigure} as solid outlines around the names of 
the soil C and N and bird and mammal body mass versus metabolic rate
datasets, in the middle row of boxes.
We did the same analyses 
with Cedar Creek data from other available years, 1996-1999 and 2007. For 1996 and 
2007, biomass and Shannon's index were independent; for 1997-1999, results 
were similar to 2000. 
<!--***DAN: cehcking the below with Shya-->
We did the same analyses with bird or mammal body mass versus BMR datasets filtered
to include only tropical species. Answers to Q1-Q3 were substantially the same.

@@

<!--***DAN: still working on this-->
Approaches used and insights gained via our work should apply equally well to 
tropical and temperate ecosystem; we believe copulas are among the tools 
that all ecologists should consider for analysis of their data in the 21st century.
There is certainly no reason to expect that non-normal copula structure and 
asymmetric tail dependence are special properties of datasets from temperate 
regions; indeed both of these features were found 
in mammal and bird body mass versus BMR data filtered to tropical
species. The Moran-effect and Liebig's law potential mechanisms that we explored
for copula structure and tail associations (Fig. \ref{MasterFigure}) also seem as 
likely to hold in tropical as in temperate systems; these are general mechanisms.
Indeed, Moran effects are widespread, and Liebig's law is a dominant paradigm,
certainly operating in a wide variety of tropical and temperate ecosystems.

>>>>

<!--***DAN: remianing to do:
DONE 1) Edit abstract
2) Go through the whole thing one more time
3) send it out to coauthors (done) and maybe get feedback, ESP FROM lAWRENCE
DONE 4)make sure to say something about tropical systems
DONE 5) maybe write to the editors saying this is coming soon?
DONE 6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.
DONE 7) change the box around the Cedar Creek data in the main fig
DONE 8) say in Discussion that you could have used non-ranks for the nonparam stats - this
is the stuff the Ecol Mongr referee talked about. Maybe this should go in Background instead?
9) maybe go back and read the Ecol Monogr reviews again
-->

@@

<!--bivariate summary-->
Table \ref{tab_bivar_summary} summarizes Q1 results 
for the bivariate datasets, with some details in the Appendices. The two 
variables were significantly related for all datasets 
(Table \ref{tab_bivar_summary}, row 1). A non-normal copula 
always emerged as the lowest-AIC copula 
(Table \ref{tab_bivar_summary}, row 2). 
The normal copula was a poor fit compared to the lowest-AIC copula 
(Table \ref{tab_bivar_summary}, rows 3-4), except for Cedar Creek, for which the AIC
difference between these fits was not meaningfully large.
Thus we answer the first part of Q1 in the affirmative for 3 of 4 of our bivariate datasets. 
Often, either lower- or upper-tail dependence statistics differed substantially from 0 
(Table \ref{tab_bivar_summary}, rows 7-8), and/or these statistics differed from each other (Table \ref{tab_bivar_summary}, row 9), 
helping to answer the second and third parts of Q1. Though again the model
selection results do not provide 
information on statistical significance of these differences, and are subject to the
caveat that, for some datasets, even the lowest-AIC copula was not an objectively good fit
(Table \ref{tab_bivar_summary}, rows 5 and 6), nonparametric results 
(Table \ref{tab_bivar_summary}, row 10) showed that tail associations
deviated significantly from what would be expected from a normal-copula
null model, and were also significantly asymmetric, except for the Cedar Creek data.
See Figs \ref{SM-fig_bmr_birds_nonparam}-\ref{SM-fig_cc2000_nonparam} for analogues to 
Fig. \ref{fig_soilCN_nonparam} for the bivariate datasets other than the soil C and N
data, see Tables \ref{SM-tab_bmr_birds_fit}-\ref{SM-tab_cc2000fit} for analogues to Table \ref{tab_soilCNfit},
and see Table \ref{SM-tab_3biv_npa_summary} for analogues to Table \ref{tab_soilCN_asym}.
The first three datasets had stronger upper- 
than lower-tail dependence, and the last dataset had the reverse, though non-significantly. Our 
generally affirmative (except for Cedar Creek), 
empirically based answer to Q1 for the bivariate datasets is 
represented in Fig. \ref{MasterFigure} as solid outlines around the names of 
the soil C and N and bird and mammal body mass versus metabolic rate
datasets, in the middle row of boxes.
We did the same analyses 
with Cedar Creek data from other available years, 1996-1999 and 2007. For 1996 and 
2007, biomass and Shannon's index were independent; for 1997-1999, results 
were similar to 2000.

@@

<!--tropical ecosystems-->
The approaches we have demonstrated should apply equally well to 
data from tropical or temperate ecosystems.
There is no reason to expect that non-normal copula structure and 
asymmetric tail dependence should be special properties of datasets from temperate 
regions. The mechanisms we have proposed of non-normal copula structure and asymmetric tail 
dependence seem equally likely to apply anywhere. For instance, the Moran effect, which underpins 
one of our proposed mechanisms (Fig. \ref{MasterFigure}B), is a standard mechanism
that occurs whenever environmental variables influence populations. And Liebig's
law and nonlinear environmental influences on ecosystems, which underpin 
another one of our proposed mechanisms (Fig. \ref{MasterFigureA}), are widely
demonstrated paradigms that apply to all biomes.
